Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest

### 2019-12-11 ms/response_to_reviews.tex

We thank the reviewer for their comments and apologize for the failure in proofreading. The reviewer should not have had to do such copy editing and I I sincerely apologize for that. We have taken heed of both reviewers' comments and fixed all specific grammar errors and fixed the issue we had with missing symbols. For brevity, we've omited response to the very simple suggested grammar fixes and simply made all requested changes. We respond to other points individually, below

\begin{quote}
line 72 - the comment about natural lightning ignitions comes from nowhere, and is a  bit of a jolt for the reader - rephrase slightly, e.g. many fires result from lightning ignitions, which are concurrent with ....
\end{quote}

TODO.

This portion was poorly written. We have rewritten this in response to reviewer 2's comment that most fires (or, at elast burned area) result from dry lightning strikes across the west. See our responses below.

@@

{\bf Responses to the Reviewer 2's comments}

\begin{quote}

My major concerns are:
1. One motivation for the study is not valid. I do not completely agree with the authors that natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event. The efficiency of individual lighting strikes in igniting a fire is affected by the moisture properties of forest fuel at the moment of lighting (Krawchuket et al. 2006, Ecology). Dry lighting that occurs without any accompanied rainfall nearby is most likely to cause wildfire because the fuel is dry enough to be ignited.
\end{quote}

Yes, this sentence was poorly written and oversimplified. We thank the reviewer
for pointing this out and we have rewritten this section. However, our badly
made point was: lighting ignitions can occur during the same general period as
precipitation. As the reviewer points out, dry strikes that find dry fuel are
more likely to lead to fires (Abatzoglou et al 2016). That the total number
lightning strikes does not well predict burned area and that ignitions occur
often from dry strikes argues for the importance of fuel moisture. Although
synoptic weather conditions that create dry and wet storms differ, the
potential for precipitation and lightning during the same overall period of
thunderstorm activity often exists (eg wet and dry convection days can occur
under overlapping conditions, Rorig and Ferguson 1999).

--  we should have said that any any precipoation that occurs during seasons when lighting is likely can make ingitions less likely

Cite:

Article{rorig1999characteristics,
  title={Characteristics of lightning and wildland fire ignition in the Pacific Northwest},
  author={Rorig, Miriam L and Ferguson, Sue A},
  journal={Journal of Applied Meteorology},
  volume={38},
  number={11},
  pages={1565--1575},
  year={1999}
}

@article{abatzoglou2016controls,
  title={Controls on interannual variability in lightning-caused fire activity in the western US},
  author={Abatzoglou, John T and Kolden, Crystal A and Balch, Jennifer K and Bradley, Bethany A},
  journal={Environmental Research Letters},
  volume={11},
  number={4},
  pages={045005},
  year={2016},
  publisher={IOP Publishing}
}

http://www.publish.csiro.au/wf/WF08117

@article{van2008temporal,
  title={Temporal and spatial distribution of lightning strikes in California in relation to large-scale weather patterns},
  author={Van Wagtendonk, Jan W and Cayan, Daniel R},
  journal={Fire Ecology},
  volume={4},
  number={1},
  pages={34--56},
  year={2008},
  publisher={Springer}
}

### 2019-12-11 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Work on edits to introduction
in response to reviews

      \item Moisture content is a strong determinant of forest fuel
    flammability and varies on small spatial scales and over short time
    periods. Empirical modeling of equilibrium dead fine fuel moisture is well
    established, but less is known about the mechanisms by which
    species-specific litter traits influence litter moisture dynamics.
    Characterizing how species’ leaf litter retain and release moisture through
    time is critical to the assessment of fuel’s availability to burn.
      \item Natural forests are often comprised of mixed stands, leading to
    mixtures of leaf litter on the surface fuelling the fires. Multiple studies
    have established the existence of non-additive effects in the flammability
    of dry litter, but we lack an understanding of moisture synergies in mixed
    litter and of moisture effects on the flammability of mixed litter.
      \item The litter from eight species of a mixed-conifer forest was
    saturated and allowed to dry to determine moisture absorption capacity and
    drying rates. Burn trials were performed in moist litter beds of single
    species and of mixtures to establish flammability response to the dry-down
    process.
      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. Species that produce
    more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
    species) have faster rates of litter drying.
      \item We examined two axes of flammability variation (spread rate and
    fuel consumption) and found that these differing moisture dynamics result
    in time since wetting having strongly differing effects on flammability
    across the major litter types typified by different tree taxa.
      \item We found that litter mixtures exhibited consistent non-additivity
    in flame spread rate in fuel consumption. This non-additivity became more
    positive with fuel drying. Litter mixtures burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.
      \item \emph{Synthesis:} In this temperate mixed conifer forest, the shade
    intolerant species are favored by the increased fire to which their fuels
    contribute Such positive feedback effects increases the potential for
    species to exert community scale effects even when relatively rare and that
    can influence community assembly and the evolution of niche-constructing
    traits.
  \end{enumerate}
\end{abstract}

\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, non-additivity

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior. Moisture acts as a heat-sink resulting in less energy
available for propagation, and the resulting water vapor will both cool the
flaming front and dilute flammable gases \citep{Albini-1976, Shafizadeh-1977}.
Environmental conditions such as temperature, relative humidity, and wind drive
the moisture content of the fuel available to burn
\citep{Kreye+Varner+etal-2018}. Variation in moisture of living fuels has
strong effects on flammability, but moisture of dead excised litter fuels can
also vary. In particular, fine dead fuels respond very readily to environmental
changes \citep{Nelson-2001}. Fuel moisture influences flammability and fire
behavior through effects on ignition, spread rate, flame height, fuel
consumption, and heat release, due to the high specific heat content of water
\citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels varies with
climate, with seasons, and with short term weather within a season. Therefore,
the rate of drying and flammability as a function of moisture content are key
parameters that influence fuel flammability throughout the fire season
\citep{Nelson-2001}.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however.  TODO REWRITE: In many litter fuel driven
systems, natural lightning ignitions are concurrent with precipitation and,
therefore, fire behavior can depend on rate of drying following a wetting
event. If species differ in their drying rate, this may influence
relatively flammability across species and over time following
wetting. Such differential moisture content across time may contribute to the
often high spatial heterogeneity of surface fires
\citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Litter packing, measured as litter-bed density or packing ratio, controls the
flammability of dry litter fuels in seasonally dry climates because higher
litter density reduces flammability as a consequence of oxygen limitation. Leaf
litter density is driven by leaf size and shape
\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
These litter particle effects on litter bed aeration, in turn, have effects on
litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
are well established. However, how moisture dynamics modulate species specific
flammability is poorly known and could be driven by by leaf traits, emergent
litter traits, or both. Species specific effects on litter flammability
are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
where species-specific litter drying rates play a role in flammability
feedbacks \citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual species traits have been recorded for western US forests as well: in mixed conifer forest in California, where the fire regime is characterized by surface fires fueled
by plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns, but
current estimates of litter flammability in mixed conifer forests are based
largely on burning dry litter.

Fuel moisture content through time is controlled by two main processes,
absorption and retention \citep{Kreye+Kobziar+etal-2013}.
Absorptive capacity represents the potential maximum fuel moisture and is
evaluated by the moisture content of litter when saturated. Differences in moisture content over time since saturation can indicate different abilities to retain
moisture. Litter particle geometry may have direct effects on maximum moisture
absorption through area-mass relationships and we also expect effects of leaf
morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption, mmoisture retention should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore effects fire behavior in two ways: directly by its effect on oxygen availability for combustion,
and indirectly by its effect on dry-down rate. Both these pathways are
affected by leaf particle traits, especially leaf size
\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
moisture retainability should favor drying, thus increasing the probability
of ignition and fire spread.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In litter driven fire, the two major
axes of variation are described by measurements related to flame spread rate
and measurements related to heat transfer [CITATIONS]  . Flame spread rate is easy to measure
in burning trials, but heat transfer can be more difficult and often proxies
such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
  Varner+Kane+etal-2015}. These different components of fire behavior have
differing ecological effects.

TODO

Species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such non-additive effects are often synergistic in dry litter fuels where
the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. TODO cite negative flammability. Zhao et al 2019]

It is possible
that similar non-additive effects may influence litter moisture content,

% especially if different leaf traits influence absorptive capacity and drying
% rate.

There is some evidence for non-additive flammability in moist fuels with
one study finding increasing non-additivity with moisture
\citep{Blauw+Wensink+etal-2015}.

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined the influence
of moisture on the flammability across these species. Experimental burn trials
of leaf litter at various moisture levels determined how time since wetting
influenced two axes of flammability characterized by flame spread rate and
total fuel consumed. Previous work described the flammability of these species
in monocultures and in mixtures for oven-dry litter
\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
flammability where fire behavior of mixtures was driven by the most flammable
species. Here, we examined if this positive non-additivity was maintained in
mixtures of varying levels of fuel moisture. The burn trials also
performed with multi-species litter mixtures to evaluate if there are non-additive effects as
observed in oven-dried litter. We tested if the non-additive effects on
reported in previous studies occur under [TODO CHECK]
ecologically-relevant moisture levels,
and whether fuel moisture effects strengthen or moderate such non-additivity.

@@

We investigated possible non-additive effects on moisture and on flammability
by testing if moisture content and flammability of mixtures were predicted by
the average of the constituent species. To choose mixtures, we first grouped
the species according to their maximum water retention and drying rate
according to estimated marginal means by species using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null with a mixed effects linear
model. For flammability parameters, we fit linear mixed effect models with the
observed-predicted flammability value as the dependent variable and used Wald
tests to test if the centered intercept was different than zero (a measure of
overall non-additivity) and if the slope with time since wetting was different
from zero (indicating non-additivity changing with drying).

@@

\subsection*{Non-additivity in  litter mixtures}

@@

There was evidence for non-additive effects in drying rate. Mixtures tended to
have lower water content than predicted from the mean of the constituent
species after wetting, but had higher water content than expected after long
drying (Wald test shows slope of mixed linear model illustrated in Fig.
\ref{fig:mixture_obs_pred_mc} less than 1, $P < 0.0001$).

@@

Flammability measures also demonstrated non-additivity in mixtures. Flame
spread rate was consistently higher than predicted (Fig
\ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
intercept for observed-predicted values greater than zero, $P$ = 0.017) and the
positive non-additivity increased with time since wetting (slope greater than
zero, $P$ = 0.005). Fuel consumption, however, showed no overall bias towards
positive or negative non-additivity, but non-additivity increased with time
since wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to
zero, $P < 0.0001$).

@@

The overall ranking of flammability in litter mixtures showed some consistency
with the results obtained for the individual species, but there were
significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
influenced by the lower bulk densities of litter mixtures containing those two
species. Most interestingly, there were patterns of non-additivity in
spread rate and in fuel consumption for the four mixtures, indicating that
positive non-additivity in flammability is not restricted to the completely dry
litter mixtures studied in previous work \citep{Magalhaes+Schwilk-2012,
  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
consumption became more positive with time since wetting and these patterns
were consistent across mixture identities. This is a slightly different result
than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
al found that non-additivity was greater in moist mixtures of heathland fuels
than in dry fuels. In their work, non-additivity was generally negative but
cases of positive non-additivity in mixtures was often a result of mixtures
meeting ignition thresholds not met by the constituent species alone [TODO: check wording]. To limit
the number of models fit, we limited our investigation to two flammability
parameters that represent two major axes of flammability variation. Therefore,
our study does not fully capture ignition effects although we do have some
results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
at times since wetting at which the species in monoculture could not.

Overall, we report a consistent pattern of increasingly positive
non-additivity in flammability with litter drying across two largely orthogonal
measures of flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018}. This result, together with previously published work,
suggests that litter flammability is prone to non-additive effects, suggesting
caution in lumping species into fuel categories. Community flammability is
dependent on and particular to its constituent species and cannot be addressed
by simply averaging the flammability of all species included. In litter-driven
fire, positive non-additivity appears to be the rule, and the relationships we
described here can form a working hypothesis and baseline expectation to adjust
fire behavior predictions in real communities: in mixtures with fuel moisture
content below 50\%, observed flame spread rates approached double the predicted
value based on monoculture trials (Fig. \ref{fig:mixture_obs_pred_spread}). In
completely dry litter from this same system, past work demonstrates observed
spread rates approached three times the expectation
\citep{Magalhaes+Schwilk-2012}.

@@

Positive non-additivity, whereby more flammable species have effects on fire
behavior disproportionate to their contribution to the litter layer, could lead
to vegetation feedbacks. In this temperate mixed conifer forest, these effects
are likely to include positive feedbacks whereby the shade intolerant species
such as the long-needled pines and the large-leaved oak, are favored by the
increased fire to which their fuels contribute \citep{Schwilk+Caprio-2011} and
similar results are reported for southeastern US forests where species
composition shifts have led to decreases in fire activity
\citep{Nowacki+Abrams-2008}. Such positive feedback effects increases the
potential for species to exert community scale effects even when relatively
rare and that can influence community assembly and the evolutionary of
niche-constructing traits \citep{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
Ongoing climate change and concurrent changes to fire regimes are linked
phenomena. Vegetation-fire feedbacks are important at a wide range of scales
from stand-level species composition shifts to the earth system
\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
demonstrate here show the potential importance of species traits and
interactions at local and stand scales where management and fire operations
meet species' biology.

>>>>

      \item Moisture content is a strong determinant of forest fuel
    flammability, however, little is known about the mechanisms by which
    species-specific litter traits influence litter moisture absorption and
    retention. Characterizing how species’ leaf litter retain moisture through
    time and how fuel from multiple species interacts are critical to the
    assessment of fuel’s availability to burn and key steps to adding species
    or taxon specific information to fire behavior prediction.

      \item To fill these gaps in knowledge, determined
    moisture absorption and retention characteristics of the litter from
    multiple tree species, investigated what litter traits influenced these
    processes, and tested for nonadditive moisture dynamics and nonadditive
    flammability in multi-species mixtures. The litter from eight tree species of a
    mixed-conifer forest was saturated and allowed to dry to determine moisture
    absorption capacity and drying rates. Burn trials were performed in moist
    litter beds of single species and of mixtures to establish flammability
    response to the dry-down process.

      \item Moisture dynamics varied across species with higher specific leaf
    area being associated with higher maximum absorption and lower litter bulk
    density being associated with faster drying rates. We examined two axes of
    flammability variation and found that these differing moisture dynamics
    result in time since wetting having strongly differing effects on
    flammability across the major litter types typified by different tree taxa.
    Litter mixtures exhibited consistent nonadditivity in flame spread rate and
    in fuel consumption. This nonadditivity became more positive with fuel
    drying.

      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
    and retention can change flammability rankings under moist compared with dry fuels.
    Litter mixtures generally burn with fire behavior more similar to that of
    the most flammable constituent species and this effect increases as fuels
    dry. [TODO more]

  \end{enumerate}
\end{abstract}

\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, nonadditivity

@@

Fuel moisture content is an important, and often driving, determinant of
surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
resulting in less energy available for propagation, and the resulting water
vapor will both cool the flaming front and dilute flammable gases
\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
temperature, relative humidity, and wind drive the moisture content of the fuel
available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
living fuels has strong effects on flammability, but moisture of dead excised
litter fuels can also vary. In particular, fine dead fuels respond very readily
to environmental changes \citep{Nelson-2001}. Fuel moisture influences
flammability and fire behavior through effects on ignition, spread rate, flame
height, fuel consumption, and heat release, due to the high specific heat
content of water \citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels
varies with climate, with seasons, and with short term weather within a season.
Therefore, the rate of drying and flammability as a function of moisture
content are key parameters that influence fuel flammability throughout the fire
season \citep{Nelson-2001}.

Flammability is a multi-dimensional characteristic of fuels
\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
different components of fire behavior. In many forests that experience surface
fire, litter properties influence fire behavior
\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
variation are described by measurements related to flame spread rate and
measurements related to heat release or heat transfer
\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
components of fire behavior have differing ecological effects with spread rate
being associated with rates of heat release and total burned area and with heat
transfer to soils or tissues predicting soil impacts of fire and the mortality
of better insulated plant tissues. A shortcoming in our current knowledge is
our integration of specific species and plant traits into fuel hazard
assessment, fire modeling, and fire effects prediction. Developing quantitative
yet robust and general ways to integrate actual plant communities into fuel
assessment will require understanding the magnitude of flammability variability
across species in space and time and parameterizing based on plant
characteristics. [TODO: get opinion on final sentences]

Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
  Fonda-2001, Scarf+Westoby-2006, Kane+Varner+etal-2008,
  Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
packing ratio, is an important control ignition and spread in models of surface
fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
Multiple studies have demonstrated that litter bed density controls the
flammability of dry litter fuels in seasonally dry climates where higher litter
density reduces flame spread rate. Across these studies, there is increasing
consensus that larger less densely packed litter beds increase oxygen
availability and burn with faster spread rates \citep{Scarf+Westoby-2006,
  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
by litter particle size and shape which is in turn influenced by leaf size, how
leaves or branches are abscised, and subsequent decomposition
\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
  Prior+Murphy+etal-2018}.

These species-specific effects are increasingly well established. However, how
moisture dynamics modulate species specific flammability is poorly known and
could be driven by by leaf traits, emergent litter traits, or by both. Species
specific effects on litter flammability are recorded for the forests of the
southeastern US \citep{Nowacki+Abrams-2008} where species-specific litter
drying rates play a role in flammability feedbacks
\citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual
species traits have been recorded for western US forests as well: in mixed
conifer forest in California, where the fire regime is characterized by surface
fires fueled by plant litter, overstory tree composition drives variation in
local fire behavior \citep{Schwilk+Caprio-2011} as a result of differential
litter flammability across tree species and as a result of synergistic
interactions in multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}.
These flammability feedbacks may help explain historical fire frequency
patterns \citep{Schwilk+Caprio-2011}, but current estimates of litter
flammability in mixed conifer forests are based largely on burning dry litter.

Fire scientists often categorize fuel by size classes that correspond to
moisture exchange rates, although such simplified categories have long been
recognized as not sufficient to describe species-specific effects on moisture
dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
dependent upon forest structure (shading and rain throughfall effects), litter
physical structure, and chemical composition of the fuel-beds
\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
models describing moisture dynamics of fine fuels are used by fire scientists
\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
usually aim to describe equilibrium conditions and at least include temperature
and relative humidity although some include precipitation and solar radiation
effects as well. Explicit models of fire behavior as a function of time since
wetting could improve fire prediction, however. Drying rates following snowmelt
or following rainfall during the fire season will influence the length of the
fire season, flammability over time, and the probability of sustained ignition,
higher spread rates and fuel consumption when ignition occurs. That species can
differ in litter drying rates has been demonstrated for at least one forest
system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
across time may contribute to the often high spatial heterogeneity of surface
fires \citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.

Fuel moisture content through time is controlled by two processes, absorption
and retention \citep{Kreye+Kobziar+etal-2013}. Absorptive capacity represents
the potential maximum fuel moisture and is evaluated by the moisture content of
litter when saturated. Differences in moisture content over time since
saturation can indicate different abilities to retain moisture. Litter particle
geometry may have direct effects on maximum moisture absorption through
area-mass relationships and we also expect effects of leaf morphology,
especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption,
moisture retention should also be influenced by leaf and litter bed traits:
surface-area-to-volume ratio and litter bulk density will influence air flow
through the litter bed. Litter packing therefore effects fire behavior in two
ways: directly by its effect on oxygen availability for combustion, and
indirectly by its effect on dry-down rate. Both these pathways are affected by
leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
Fuel-beds composed of larger leaves with lower moisture retainability should
favor drying, thus increasing the probability of ignition and fire spread.

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+Cornwell+etal-2016}. It is possible that
similar nonadditive effects may influence litter moisture content directly and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}.

[TODO: reseearch question here]

We investigated species specific litter moisture dynamics across eight tree
species in a temperate fire-prone forest in California. We determined the
maximum water retention and rate of moisture loss and examined how these vary
in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined
the influence of moisture on the flammability across these species.
Experimental burn trials of leaf litter at various moisture levels determined
how time since wetting influenced two axes of flammability characterized by
flame spread rate and total fuel consumed. Previous work described the
flammability of these species in monocultures and in mixtures for oven-dry
litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
nonadditivity of flammability where fire behavior of mixtures was driven by the
most flammable species. Here, we examined if this positive nonadditivity was
maintained in mixtures of varying levels of fuel moisture. The burn trials also
performed with multi-species litter mixtures to evaluate if there are
nonadditive effects as observed in oven-dried litter. We tested if the
nonadditive effects on reported in previous studies occur under [TODO CHECK]
ecologically-relevant moisture levels, and whether fuel moisture effects
strengthen or moderate such nonadditivity.

@@

We investigated possible nonadditive effects on moisture and on flammability by
testing if moisture content and flammability of mixtures were predicted by the
average of the constituent species. To choose mixtures, we first grouped the
species according to their maximum water retention and drying rate according to
estimated marginal means by species using the R package `emmeans'
\citep{Lenth-2019}. This resulted in four groups of species and we then
selected one representative species from each litter type group and produced
all four possible three-species litter mixtures using those four species. For
each mixture and time since wetting, we predicted moisture content based on the
mean content of the three constituent species at that time since wetting. We
predicted flammability parameters based on the average flammability measures
from individual species trials at each time since wetting and then calculated a
predicted spread rate and fuel consumption as the average of these three values
for the mixture. The expectation from the null model was that the difference
between observed and predicted values of flammability will be zero. We tested
if there were significant departures from the null with a mixed effects linear
model. For flammability parameters, we fit linear mixed effect models with the
observed-predicted flammability value as the dependent variable and used Wald
tests to test if the centered intercept was different than zero (a measure of
overall nonadditivity) and if the slope with time since wetting was different
from zero (indicating nonadditivity changing with drying).

@@

\subsection*{Nonadditivity in  litter mixtures}

@@

There was evidence for nonadditive effects in drying rate. Mixtures tended to
have lower water content than predicted from the mean of the constituent
species after wetting, but had higher water content than expected after long
drying (Wald test shows slope of mixed linear model illustrated in Fig.
\ref{fig:mixture_obs_pred_mc} less than 1, $P < 0.0001$).

@@

Flammability measures also demonstrated nonadditivity in mixtures. Flame
spread rate was consistently higher than predicted (Fig
\ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
intercept for observed-predicted values greater than zero, $P$ = 0.017) and the
positive nonadditivity increased with time since wetting (slope greater than
zero, $P$ = 0.005). Fuel consumption, however, showed no overall bias towards
positive or negative nonadditivity, but nonadditivity increased with time
since wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to
zero, $P < 0.0001$).

@@

The overall ranking of flammability in litter mixtures showed some consistency
with the results obtained for the individual species, but there were
significant nonadditive effects. Mixtures containing \emph{Q. kelloggii} and
\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
influenced by the lower bulk densities of litter mixtures containing those two
species. Most interestingly, there were patterns of nonadditivity in spread
rate and in fuel consumption for the four mixtures, indicating that positive
nonadditivity in flammability is not restricted to the completely dry litter
mixtures studied in previous work \citep{Magalhaes+Schwilk-2012,
  VanAltena+Logtestjin+etal-2012}. Nonadditivity in both spread rate and fuel
consumption became more positive with time since wetting and these patterns
were consistent across mixture identities. This is a slightly different result
than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
al found that nonadditivity was greater in moist mixtures of heathland fuels
than in dry fuels. In their work, nonadditivity was generally negative but
cases of positive nonadditivity in mixtures was often a result of mixtures
meeting ignition thresholds not met by the constituent species alone [TODO:
check wording]. To limit the number of models fit, we limited our investigation
to two flammability parameters that represent two major axes of flammability
variation. Therefore, our study does not fully capture ignition effects
although we do have some results consistent with theirs: mixtures with \emph{Q.
  kellogii} could ignite at times since wetting at which the species in
monoculture could not.

Overall, we report a consistent pattern of increasingly positive nonadditivity
in flammability with litter drying across two largely orthogonal measures of
flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017,
  Prior+Murphy+etal02018}. This result, together with previously published
work, suggests that litter flammability is prone to nonadditive effects,
suggesting caution in lumping species into fuel categories. Community
flammability is dependent on and particular to its constituent species and
cannot be addressed by simply averaging the flammability of all species
included. In litter-driven fire, positive nonadditivity appears to be the rule,
and the relationships we described here can form a working hypothesis and
baseline expectation to adjust fire behavior predictions in real communities:
in mixtures with fuel moisture content below 50\%, observed flame spread rates
approached double the predicted value based on monoculture trials (Fig.
\ref{fig:mixture_obs_pred_spread}). In completely dry litter from this same
system, past work demonstrates observed spread rates approached three times the
expectation \citep{Magalhaes+Schwilk-2012}.

@@

Positive nonadditivity, whereby more flammable species have effects on fire
behavior disproportionate to their contribution to the litter layer, could lead
to vegetation feedbacks. In this temperate mixed conifer forest, these effects
are likely to include positive feedbacks whereby the shade intolerant species
such as the long-needled pines and the large-leaved oak, are favored by the
increased fire to which their fuels contribute \citep{Schwilk+Caprio-2011} and
similar results are reported for southeastern US forests where species
composition shifts have led to decreases in fire activity
\citep{Nowacki+Abrams-2008}. Such positive feedback effects increases the
potential for species to exert community scale effects even when relatively
rare and that can influence community assembly and the evolutionary of
niche-constructing traits \citep{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
Ongoing climate change and concurrent changes to fire regimes are linked
phenomena. Vegetation-fire feedbacks are important at a wide range of scales
from stand-level species composition shifts to the earth system
\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
demonstrate here show the potential importance of species traits and
interactions at local and stand scales where management and fire operations
meet species' biology.
